---
title: "Data Exploration in R"
output: html_document
author: Deepshikha Adhikari
date: 03/30/2019
---
PART 1: DATA EXPLORATION (30 min)

Install Packages
```{r, warning=FALSE}
#install.packages('nycflights13')
#install.packages('lubridate')
#install.packages('dplyr')
#install.packages('skimr')
library(nycflights13)
library(skimr)
library(lubridate)
library(dplyr)
```

Before getting started, use the right file 
```{r}
filepath_to_directory <- '~/Documents/datafest/2019/R'
filepath_to_data <- '~/Documents/datafest/2019/data/'
setwd(filepath_to_directory)
```


Load Data
```{r}
#Load single file at a time:
city_attributes <- read.csv(file = paste(filepath_to_data,'city_attributes.csv', sep = ""))
city_attributes <- read.csv(file = '~/Documents/Data_Exploration/data/city_attributes.csv')

#Load all files altogether
temp = list.files(path = filepath_to_data, pattern = "*.csv")
for (i in 1:length(temp)) 
  assign(temp[i], read.csv(file = paste(filepath_to_data,temp[i], sep = "")))

```


Let's pick one dataset
```{r}
#Dimension of the dataset
humidity.csv %>% 
  dim()

#Names of the columns
humidity.csv %>% names()

#Top n of the dataset
humidity.csv %>% head(n = 5) 

#Bottom n of the dataset
humidity.csv %>% tail(n = 5) 

#Random n rows
sample_n(humidity.csv, 5)

```

Sample the dataset.
```{r}
#Pop quiz: When you might wanna use it?
set.seed(100)
smp_size <- floor(0.8 * nrow(humidity.csv))
train_ind <- sample(seq_len(nrow(humidity.csv)), size = smp_size)

train <- humidity.csv[train_ind, ]
test <- humidity.csv[-train_ind, ]

```


How to sort the data?
```{r}
#Sort the data
humidity.csv %>% head(n = 5) %>% select(datetime)

#Decreasing Order
humidity.csv %>% arrange(desc(datetime)) %>% select(datetime) %>% head()

#Increasing Order 
humidity.csv %>% arrange(datetime) %>% select(datetime) %>% head()

#Increasing Order 
pressure.csv %>% arrange(Atlanta, Portland) %>% select(Atlanta,Portland) %>% head(n = 5)
```


The most amazing function for Exploration in R
-Data Type?
-Missing vals
- Mean/Max/Unique 
```{r}
skimr::skim(humidity.csv)
```


Interested in a single variables:
```{r}
#Min/Max
humidity.csv %>% 
  select(Seattle) %>% 
  min(na.rm = TRUE)

#Frequency Table
humidity.csv %>% 
  select(Seattle) %>% 
  table()
#OR
table(humidity.csv$Seattle)

#Total Unique items
length(unique(humidity.csv$Seattle))

#Data type 
class(humidity.csv)
class(humidity.csv$datetime)

#histogram
hist(humidity.csv$Seattle)

```


Dealing with missing data
```{r}
#Missing 
sum(is.na(humidity.csv$Seattle))

#front-fill


#back-fill

#data imputer with mean


#data imputer with median

```


Getting rid of duplicate Dataset:
```{r}
pressure.csv %>% select(Atlanta) %>% distinct()
```

Changing datatype
```{r}

```

Count average sum
```{r}

```


============================================================================================================
PART 2: LUBRIDATE(20 min)
https://r4ds.had.co.nz/
```{r}
skimr::skim(flights)
names(flights)
```

Parsing dates
========================================================
year, month, day
```{r}
lubridate::ymd("20100604")
lubridate::mdy("06-04-2011")
lubridate::dmy("04/06/2011")
lubridate::ymd("")
```

Parsing datetime
========================================================
year, month, day, hour, minute, second, **datetime**
```{r}
ymd_hms("2015-09-09 14:00:00")
#as_datetime('2015-09-03 16:37:00Z')

```

Parsing and Formats (DateTime)
========================================================
```{r}
johns_bday <- ymd_hms("1991-01-2000:20:00")
date(johns_bday)
hour(johns_bday)
```

Flights!
========================================================
take a look at the flights data set
```{r}
cat(names(flights))
```

Mutate a Date!
========================================================
incremental: true
```{r}
flights <- flights %>% mutate(date = ymd(paste0(year, "-", month, "-", day)))
```


```{r}
flights <- flights %>% mutate(date_time = ymd_hm(paste0(year, "-", month, "-", day, " ", hour, ":", minute)))
```

```{r}
flights %>% select(date_time)
```

Sort by Date
========================================================
```{r}
flights %>% arrange(date) %>% select(date) %>% head()
```

Accessor Functions
========================================================
what day of the week was Dana born on?
```{r}
dees_bday <- ymd("1991-09-30")
month(danas_bday)
wday(danas_bday)
```

First Flight of each Month (Question)
========================================================
aggregator functions!
```{r}
flights %>% arrange(date_time) %>% group_by(month(date_time)) %>% 
    filter(row_number() == 1) %>% ungroup() %>% select(date_time, tailnum)
```

Time Series
========================================================
get all flights of a certain plane, "N14228"
```{r}
flights %>% filter(tailnum == 'N14228') %>% arrange(date_time) %>% select(date_time, tailnum)
```

what day of the week was N0EGMQ's 100th flight?
```{r}
flights %>% filter(tailnum == 'N0EGMQ') %>% 
    arrange(date_time) %>% filter(row_number() == 100) %>%
    select(date_time) %>% mutate(wday(date_time))
```

Today!
========================================================
what if we want the date today?
```{r}
date <- today()
date
```

Intervals!
========================================================
how old is John?
```{r}
johns_bday <- ymd("1991-01-20")
today() - johns_bday
```

Durations!
========================================================
how old is John?
```{r}
johns_bday <- ymd("1991-01-20")
johns_duration <- as.duration(today() - johns_bday)
johns_duration
```

Durations!
========================================================
how old is John... in weeks?
```{r}
johns_bday <- ymd("1991-01-20")
johns_duration <- as.duration(today() - johns_bday)
johns_duration / dweeks(1)
```




